---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bnns: Bayesian Neural Networks for Regression and Classification

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![Codecov test coverage](https://codecov.io/gh/swarnendu-stat/bnns/graph/badge.svg)](https://app.codecov.io/gh/swarnendu-stat/bnns)
<!-- badges: end -->

The `bnns` package provides tools to fit Bayesian Neural Networks (BNNs) for regression and classification problems. It is designed to be flexible, supporting various network architectures, activation functions, and output types, making it suitable for both simple and complex data analysis tasks.

## Features

- Support for multi-layer neural networks with customizable architecture.
- Choice of activation functions (e.g., sigmoid, ReLU, tanh).
- Outputs for regression (continuous response) and classification (binary and multiclass).
- Bayesian inference, providing posterior distributions for predictions and parameters.
- Applications in domains such as clinical trials, predictive modeling, and more.

## Installation

To install the `bnns` package from GitHub, use the following:

```{r}
# Install devtools if not already installed
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

# Install bnns
devtools::install_github("swarnendu-stat/bnns")
```

## Getting Started

### 1. Simulate Data

Below is an example of how to simulate data for regression:

```{r}
set.seed(123)
n <- 100
x <- matrix(runif(n * 2), ncol = 2)
y <- x[, 1] + x[, 2]^2 + rnorm(n, sd = 0.1)
```

### 2. Fit a BNN Model

To fit a Bayesian Neural Network:

```{r}
library(bnns)

model <- bnns(x, y, L = 2, nodes = c(10, 8), act_fn = c(2, 3), out_act_fn = 1,
              iter = 2e2, warmup = 1e2, chains = 1, seed = 123)
```

### 3. Model Summary

Summarize the fitted model:

```{r}
summary(model)
```

### 4. Predictions

Make predictions using the trained model:

```{r}
pred <- predict(model, newdata = x)
```

### 5. Visualization

Visualize true vs predicted values for regression:

```{r}
plot(y, rowMeans(pred), main = "True vs Predicted", xlab = "True Values", ylab = "Predicted Values")
abline(0, 1, col = "red")
```

## Applications

### Regression Example

Use `bnns` for regression analysis to model continuous outcomes, such as predicting patient biomarkers in clinical trials.

### Classification Example

For binary or multiclass classification, set the `out_act_fn` to `2` (binary) or `3` (multiclass). For example:

```{r}
# Simulate binary classification data
train_x <- matrix(runif(200), ncol = 2)
train_y <- ifelse(train_x[, 1] + train_x[, 2] > 1, 1, 0)

# Fit a binary classification BNN
model <- bnns(train_x, train_y, L = 2, nodes = c(16, 8), act_fn = c(3, 2), out_act_fn = 2, iter = 2e2, warmup = 1e2, chains = 1)
```

### Clinical Trial Applications

Explore posterior probabilities to estimate treatment effects or success probabilities in clinical trials. For example, calculate the posterior probability of achieving a clinically meaningful outcome in a given population.

## Documentation

- Detailed vignettes are available to guide through various applications of the package.
- See `help(bnns)` for more information about the `bnns` function and its arguments.

## Contributing

Contributions are welcome! Please raise issues or submit pull requests on [GitHub](https://github.com/swarnendu-stat/bnns).

## License

This package is licensed under the Apache License. See `LICENSE` for details.
