---
title: "Using Bayesian Neural Networks in Clinical Trials"
author: "Swarnendu Chatterjee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Bayesian Neural Networks in Clinical Trials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(bnns)
```

## Introduction

Bayesian Neural Networks (BNNs) offer a robust framework for prediction in clinical trials by providing posterior distributions of predictions. This allows for probabilistic reasoning, such as computing the probability that a treatment achieves a certain efficacy threshold or proportion of success.

In this vignette, we:
1. Illustrate data preparation for a clinical trial setting.
2. Fit a BNN to simulate clinical trial outcomes.
3. Leverage posterior distributions for decision-making, such as calculating posterior probabilities of treatment success.

---

## 1. Data Preparation

Consider a hypothetical clinical trial comparing the efficacy of a new treatment against a placebo. The response variable is binary, representing treatment success (1) or failure (0).

### Simulating Data

```{r}
set.seed(123)

# Simulate predictor variables (e.g., patient covariates)
n_subjects <- 100
Age <- runif(n_subjects, 18, 50) # Age in years
Dose <- runif(n_subjects, 10, 100) # Dose levels
Severity <- runif(n_subjects, 1, 10) # Baseline severity (arbitrary scale)

# Define true probabilities using a nonlinear function
beta_0 <- 1
beta_1 <- 0.3
beta_2 <- -0.1
beta_3 <- -0.02
beta_4 <- 0.005

logit_p <- beta_0 + beta_1 * Dose + beta_2 * log(Severity) +
  beta_3 * Age^2 + beta_4 * (Age * Dose)
p_success <- 1 / (1 + exp(-logit_p)) # Sigmoid transformation

# Simulate binary outcomes
Success <- rbinom(n_subjects, size = 1, prob = p_success)

trial_data <- cbind.data.frame(Success, Age, Dose, Severity)

# Split into training and testing
train_idx <- sample(seq_len(n_subjects), size = 0.8 * n_subjects)
training_data <- trial_data[train_idx, ]
test_data <- trial_data[-train_idx, ]
```

---

## 2. Fitting a Bayesian Neural Network

Fit a BNN to the simulated data. We use a binary classification model with a logistic sigmoid activation for the output layer.

```{r message=FALSE, warning=FALSE, echo = TRUE, results = 'hide'}
# Fit a BNN
model <- bnns(
  formula = Success ~ -1 + .,
  data = training_data,
  L = 2, # Number of hidden layers
  nodes = c(16, 8), # Nodes per layer
  act_fn = c(2, 2), # Activation functions for hidden layers
  out_act_fn = 2, # Output activation: logistic sigmoid
  iter = 2e2, # Bayesian sampling iterations
  warmup = 1e2, # Warmup iterations
  chains = 1 # Number of MCMC chains
)
```

---

## 3. Posterior Predictions

### Generating Predictions with Uncertainty

The posterior distribution of predictions allows us to compute not just point estimates but also probabilistic metrics.

```{r}
# Generate posterior predictions for the test set
posterior_preds <- predict(model, subset(test_data, select = -Success))
head(posterior_preds) # Each row corresponds to a subject, and columns are MCMC samples
```

Each entry in `posterior_preds` represents the predicted probability of success from a single posterior sample.

---

## 4. Posterior Probability of Treatment Success

### Threshold-Based Decision-Making

Suppose we define treatment success as a predicted probability ≥ 0.6. We can compute the posterior probability that this threshold is met for each subject.

```{r}
# Compute posterior probabilities of success (p_hat ≥ 0.6)
success_threshold <- 0.6
posterior_probs_success <- rowMeans(posterior_preds >= success_threshold)
head(posterior_probs_success)
```

### Proportion of Subjects Likely to Achieve Success

Next, we calculate the posterior probability that a certain proportion of subjects in the treatment group achieve success.

```{r}
# Define success proportion threshold
prop_success_threshold <- 0.7

# Simulate posterior proportion of success
posterior_success_proportion <- colMeans(posterior_preds >= success_threshold)

# Posterior probability that ≥ 70% of subjects achieve success
posterior_prob_high_success <- mean(posterior_success_proportion >= prop_success_threshold)
posterior_prob_high_success
```

---

## 5. Visualizing Posterior Insights

### Posterior Predictive Distribution

```{r}
library(ggplot2)

# Plot posterior probabilities of success for individual subjects
ggplot(data.frame(Subject = seq_len(nrow(test_data)), Prob = posterior_probs_success), aes(x = Subject, y = Prob)) +
  geom_bar(stat = "identity", fill = "blue") +
  geom_hline(yintercept = success_threshold, color = "red", linetype = "dashed") +
  labs(
    title = "Posterior Probability of Treatment Success",
    x = "Subject",
    y = "Posterior Probability"
  )
```

### Distribution of Success Proportion

```{r}
# Histogram of posterior success proportions
ggplot(data.frame(SuccessProp = posterior_success_proportion), aes(x = SuccessProp)) +
  geom_histogram(fill = "green", bins = 20) +
  geom_vline(xintercept = prop_success_threshold, color = "red", linetype = "dashed") +
  labs(
    title = "Posterior Distribution of Success Proportion",
    x = "Proportion of Subjects Achieving Success",
    y = "Frequency"
  )
```

---

## 6. Clinical Trial Decision-Making

### Bayesian Probability Threshold

The posterior probability can guide decision-making. For example:
- If `posterior_prob_high_success > 0.9`, consider the treatment effective.
- If `posterior_prob_high_success < 0.1`, consider the treatment ineffective.
- Otherwise, collect more data or refine the model.

---

## 7. Conclusion

The `bnns` package empowers clinical trial analysts to leverage Bayesian Neural Networks for predictive modeling and decision-making. By utilizing posterior distributions, we can:
- Quantify uncertainty in predictions.
- Make informed decisions about treatment efficacy.
- Evaluate trial outcomes based on predefined success criteria.

This probabilistic framework is particularly valuable in scenarios where uncertainty plays a critical role in decision-making, such as early-phase clinical trials.

```
