---
title: "Benchmarking bnns"
---

```{r, include = FALSE}
has_pkgs <- requireNamespace("mlbench", quietly = TRUE) && 
            requireNamespace("tidymodels", quietly = TRUE) && 
            requireNamespace("ranger", quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_pkgs
)
```

```{r setup}
# Ensuring required packages are installed before proceeding
stopifnot("mlbench not installed" = requireNamespace("mlbench", quietly = TRUE))
stopifnot("rsample not installed" = requireNamespace("rsample", quietly = TRUE))
stopifnot("ranger not installed" = requireNamespace("ranger", quietly = TRUE))
stopifnot("parsnip not installed" = requireNamespace("parsnip", quietly = TRUE))

# Loading the necessary libraries
library(bnns)    # For Bayesian Neural Networks
library(parsnip) # For unified modeling interface
library(mlbench) # For benchmark datasets
library(rsample) # For data splitting
set.seed(123)    # Setting seed for reproducibility
```

## Introduction

This article demonstrates the performance of the `bnns` package on three datasets from the `mlbench` package:  

- **Regression**: `mlbench.friedman1` dataset  
- **Binary Classification**: `mlbench.spirals` dataset  
- **Multi-class Classification**: `mlbench.waveform` dataset  

We utilize the `parsnip` framework from the `tidymodels` ecosystem to provide a unified interface for model training and prediction. For each dataset, we:

1. Prepare the data for training and testing.
2. Build a Bayesian Neural Network using the `bnns` engine via `parsnip::mlp()`.
3. Evaluate the model's predictive performance.
4. To compare, show the performance of the random forest algorithm using the `ranger` engine.

---

## Regression: Friedman1 Dataset

### Dataset Description
The dataset generated with `mlbench.friedman1` is the regression problem Friedman 1 as described in Friedman (1991) and Breiman (1996). Inputs are 10 independent variables uniformly distributed on the interval [0,1], only 5 out of these 10 are actually used. Outputs are created according to the formula

$$
y = 10 \sin(\pi x_1 x_2) + 20 (x_3 - 0.5)^2 + 10 x_4 + 5 x_5 + \epsilon
$$

$\epsilon \sim N(0, 1).$

```{r regression-data}
# Generating the Friedman1 dataset
friedman1_data <- mlbench.friedman1(n = 100, sd = 1)
friedman1_df <- cbind.data.frame(y = friedman1_data$y, friedman1_data$x)
colnames(friedman1_df) <- c("y", paste0("x", 1:10))

# Splitting the data into training (80%) and testing (20%) sets
friedman1_split <- initial_split(friedman1_df, prop = 0.8)
friedman1_train <- training(friedman1_split)  # Training data
friedman1_test <- testing(friedman1_split)   # Testing data
```

### Model Training
```{r regression-model}
# Specifying a Bayesian Neural Network with a single hidden layer and 4 nodes
friedman1_bnn_spec <- mlp(
  mode = "regression",
  hidden_units = 4,
  epochs = 1000,
  activation = "softplus"
) |> 
  set_engine("bnns", warmup = 200)

# Training the model
friedman1_bnn_fit <- friedman1_bnn_spec |> 
  fit(y ~ ., data = friedman1_train)
```

### Model Evaluation
```{r regression-evaluation}
# Making predictions on the test set and evaluating model performance
friedman1_bnn_pred <- predict(friedman1_bnn_fit, new_data = friedman1_test)
bnns::measure_cont(friedman1_test$y, friedman1_bnn_pred$.pred)  # Measures like RMSE, MAE
```

### Model Comparison
```{r regression-comparison}
# Training a random forest model for comparison
friedman1_rf_spec <- rand_forest(mode = "regression") |> 
  set_engine("ranger")

friedman1_rf_fit <- friedman1_rf_spec |> 
  fit(y ~ ., data = friedman1_train)

# Making predictions with random forest and evaluating performance
friedman1_rf_pred <- predict(friedman1_rf_fit, new_data = friedman1_test)
measure_cont(friedman1_test$y, friedman1_rf_pred$.pred)
```

---

## Binary Classification: Spirals Dataset

### Dataset Description
The dataset generated with the `mlbench.spirals` consists of points on two entangled spirals. If `sd>0`, then Gaussian noise is added to each data point.

```{r binary-data}
# Generating the Spirals dataset with Gaussian noise
spirals_data <- mlbench.spirals(100, 1.5, 0.05)
spirals_df <- cbind.data.frame(y = spirals_data$classes, spirals_data$x)
colnames(spirals_df) <- c("y", "x1", "x2")

# Splitting the data into training and testing sets (stratified by class)
spirals_split <- initial_split(spirals_df, prop = 0.8, strata = "y")
spirals_train <- training(spirals_split)  # Training data
spirals_test <- testing(spirals_split)   # Testing data
```

### Model Training
```{r binary-model}
# Training a Bayesian Neural Network with three hidden layers
spirals_bnn_spec <- mlp(
  mode = "classification", 
  hidden_units = c(64, 64, 16),
  epochs = 1000,
  activation = c("tanh", "relu", "relu")
) |> 
  set_engine("bnns", L = 3, warmup = 200)

spirals_bnn_fit <- spirals_bnn_spec |> 
  fit(y ~ ., data = spirals_train)
```

### Model Evaluation
```{r binary-evaluation}
# Making predictions and calculating binary classification metrics (e.g., AUC)
spirals_bnn_pred <- predict(spirals_bnn_fit, new_data = spirals_test, type = "prob")
# Convert factor to numeric 0/1 for measure_bin
measure_bin(as.numeric(spirals_test$y) - 1, spirals_bnn_pred$.pred_2)
```

### Model Comparison
```{r binary-comparison}
# Training a random forest model for comparison
spirals_rf_spec <- rand_forest(mode = "classification") |> 
  set_engine("ranger")

spirals_rf_fit <- spirals_rf_spec |> 
  fit(y ~ ., data = spirals_train)

# Evaluating the random forest model
spirals_rf_pred <- predict(spirals_rf_fit, new_data = spirals_test, type = "prob")
measure_bin(as.numeric(spirals_test$y) - 1, spirals_rf_pred$.pred_2)
```

---

## Multi-class Classification: Waveform Dataset

### Dataset Description
The dataset generated with `mlbench.waveform` consists of 21 attributes with continuous values and a variable showing the 3 classes (33% for each of 3 classes). Each class is generated from a combination of 2 of 3 "base" waves.

```{r multi-data}
# Generating the Waveform dataset
waveform_data <- mlbench.waveform(100)
waveform_df <- cbind.data.frame(y = waveform_data$classes, waveform_data$x)
colnames(waveform_df) <- c("y", paste0("x", 1:21))

# Splitting the data into training and testing sets (stratified by class)
waveform_split <- initial_split(waveform_df, prop = 0.8, strata = "y")
waveform_train <- training(waveform_split)  # Training data
waveform_test <- testing(waveform_split)   # Testing data
```

### Model Training
```{r multi-model}
# Training a Bayesian Neural Network with two hidden layers
waveform_bnn_spec <- mlp(
  mode = "classification", 
  hidden_units = c(2, 2),
  epochs = 1000,
  activation = c("sigmoid", "softplus")
) |> 
  set_engine("bnns", L = 2, warmup = 200)

waveform_bnn_fit <- waveform_bnn_spec |> 
  fit(y ~ ., data = waveform_train)
```

### Model Evaluation
```{r multi-evaluation}
# Making predictions and evaluating multi-class classification metrics
waveform_bnn_pred <- predict(waveform_bnn_fit, new_data = waveform_test, type = "prob")
measure_cat(waveform_test$y, as.matrix(waveform_bnn_pred))
```

### Model Comparison
```{r multi-comparison}
# Training a random forest model with probability outputs for comparison
waveform_rf_spec <- rand_forest(mode = "classification") |> 
  set_engine("ranger")

waveform_rf_fit <- waveform_rf_spec |> 
  fit(y ~ ., data = waveform_train)

# Evaluating the random forest model
waveform_rf_pred <- predict(waveform_rf_fit, new_data = waveform_test, type = "prob")
measure_cat(waveform_test$y, as.matrix(waveform_rf_pred))
```

---

## Summary

The `bnns` package showcases strong predictive performance across regression, binary classification, and multi-class classification tasks. In addition to accurate predictions, it provides posterior distributions, enabling:

1. Uncertainty Quantification: Offers insights into the confidence of predictions, crucial for high-stakes applications like clinical trials and finance.
2. Probabilistic Decision-Making: Facilitates decisions under uncertainty by integrating Bayesian principles.
3. Model Comparisons: Demonstrates comparable performance to the ranger package, with the added advantage of interpretability through Bayesian inference.

Overall, `bnns` is a powerful tool for tasks requiring both predictive accuracy and interpretability, making it suitable for various domains, and its integration with the `tidymodels` framework via `parsnip` makes it easy to integrate into modern R modeling workflows.
