---
title: "Common Application Areas of bnns"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
stopifnot("mlbench not installed" = requireNamespace("mlbench", quietly = TRUE))
stopifnot("rsample not installed" = requireNamespace("rsample", quietly = TRUE))
library(bnns)
library(mlbench)
library(rsample)
set.seed(123)
```

## Introduction

This article demonstrates the use of the `bnns` package on three datasets from the `mlbench` package:  

- **Regression**: `BostonHousing` dataset  
- **Binary Classification**: `PimaIndiansDiabetes` dataset  
- **Multi-class Classification**: `Glass` dataset  

For each dataset, we:
1. Prepare the data for training and testing.
2. Build a Bayesian Neural Network using the `bnns` package.
3. Evaluate the model's predictive performance.

---

## Regression: BostonHousing Dataset

### Dataset Description
The `BostonHousing` dataset contains information on housing prices in Boston, with features like crime rate, average number of rooms, and more.

```{r regression-data}
data(BostonHousing)
BH_data <- BostonHousing
# Splitting data into training and testing sets
BH_split <- initial_split(BH_data, prop = 0.8)
BH_train <- training(BH_split)
BH_test <- testing(BH_split)
```

### Model Training
```{r regression-model}
model_reg <- bnns(
  medv ~ -1 + .,
  data = BH_train, L = 2, out_act_fn = 1,
  iter = 1e3, warmup = 2e2, chains = 2, cores = 2
)
```

### Model Evaluation
```{r regression-evaluation}
BH_pred <- predict(model_reg, newdata = BH_test)
measure_cont(BH_test$medv, BH_pred)
```

---

## Binary Classification: PimaIndiansDiabetes Dataset

### Dataset Description
The `PimaIndiansDiabetes` dataset contains features related to health status for predicting the presence of diabetes.

```{r binary-data}
data(PimaIndiansDiabetes)
PID_data <- PimaIndiansDiabetes |>
  transform(diabetes = ifelse(diabetes == "pos", 1, 0))
# Splitting data into training and testing sets
PID_split <- initial_split(PID_data, prop = 0.8, strata = "diabetes")
PID_train <- training(PID_split)
PID_test <- testing(PID_split)
```

### Model Training
```{r binary-model}
model_bin <- bnns(
  diabetes ~ -1 + .,
  data = PID_train, L = 2,
  out_act_fn = 2, iter = 1e3, warmup = 2e2, chains = 2, cores = 2
)
```

### Model Evaluation
```{r binary-evaluation}
PID_pred <- predict(model_bin, newdata = PID_test)
PID_measure <- measure_bin(PID_test$diabetes, PID_pred)
PID_measure
plot(PID_measure$ROC)
```

---

## Multi-class Classification: Glass Dataset

### Dataset Description
The `Glass` dataset contains features to classify glass types.

```{r multi-data}
data(Glass)
Glass_data <- Glass

# Splitting data into training and testing sets
Glass_split <- initial_split(Glass_data, prop = 0.8, strata = "Type")
Glass_train <- training(Glass_split)
Glass_test <- testing(Glass_split)
```

### Model Training
```{r multi-model}
model_multi <- bnns(
  Type ~ -1 + .,
  data = Glass_train, L = 2,
  out_act_fn = 3, iter = 1e3, warmup = 2e2, chains = 2, cores = 2
)
```

### Model Evaluation
```{r multi-evaluation}
Glass_pred <- predict(model_multi, newdata = Glass_test)
measure_cat(Glass_test$Type, Glass_pred)
```

---

## Summary

The performance of the `bnns` package demonstrates its flexibility across various machine learning tasks. It provides posterior distributions of predictions, which can be used for uncertainty quantification and probabilistic decision-making.
